Positivity-Preserving Adaptive Runge--Kutta Methods

### 2020-05-04 nusslein_vsrp/latex/overview.tex
WIP: further revision

We say an initial value problem
\begin{align} \label{ode}
    u'(t) & = f(t,u) \\
    u(0) & = u_0
\end{align}
where $u: [0,T] \to \R^m$ is positive if
\begin{align} \label{continuous-positivity}
    u(0)\ge 0 \implies u(t) \ge 0 \text{ for all } t \in [0,T].
\end{align}
A sufficient condition for positivity of \eqref{ode} is
\begin{equation} \label{eq:condition_ODE_pos}
u_i=0 \implies f_i(t,u_1,\cdots,u_i,\cdots,u_n) \geq 0 \quad \forall {u \geq 0,}\; \ \  \forall {t\in[0,T]}.
\end{equation}
For such ODEs, the backward Euler method is guaranteed to preserve positivity
under any step size, while the forward Euler will preserve positivity for small enough $\dt$ \cite{hundsdorfer_numerical_2003}.
Any RKM (or in fact any general linear method) that is unconditionally positivity preserving for all positive ODEs
must have order $\le 1$ \cite{bolley_conservation_1978}.
For any higher order method, we expect positivity only under some restriction
on the time step size.

@@

The idea of using different weights within an RKM is not new; for instance it is
the basis of error approximation using embedded RK pairs \cite{hairer_solving_1993}.
The idea of adapting the weights after calculating the stage values has also been used,
for instance in \cite{ketcheson_spatially_2013}.  In this case it is used to
adapt the properties of the time integrator for a method of lines solution of a
PDE.  Another class of methods that adapt the weights at the end of an RK step
are the relaxation Runge--Kutta (RRK) methods.  In these, the weights are
scaled by a scalar relaxation parameter in order to guarantee conservation or
monotonicity of a desired functional; e.g. to conserve or dissipate energy or
entropy
\cite{ketcheson_relaxation_2019,ranocha_relaxation_2019,ranocha2020general}.

Our means to ensure positivity can be interpreted as a projection
approach, where the numerical solution is adapted to satisfy the
positivity constraint at the end of each step. In contrast to
simple orthogonal projection, which has also been proposed to deal
with positivity constraints \cite{shampine1986conservation},
our approach preserves all linear invariants of the given ODE. These
invariants can be very important. e.g.\ the total mass for a transport
problem or in reaction systems. Preservation of linear
invariants has been shown to be an important advantage of RRK methods
over orthogonal projection methods
\cite{ranocha2020relaxationHamiltonian}. Of course, it is also
possible to enforce the preservation of linear invariants in projection
methods, but the invariants have to be known explicitly
\cite{sandu2001positive}.

@@

When computing the solution of an ODE $u ' = f(t,u) $ using an RKM with $s$ stages and the Butcher tableau
\begin{align}
\renewcommand{\arraystretch}{1.2}
\begin{array}{c|c}
c &  A \\
\hline
 & b^T\\
\end{array}
\end{align}
the stage values are computed according to
\begin{equation}\label{eq:stagevalues}
y_j =  u^n + \dt \sum_{k = 1}^{s} a_{jk} f(t^n + \dt c_k,y_k),  \quad j = 1,\cdots,s.
\end{equation}
Based on these values, the next solution $u^{n+1}$ is computed as
\begin{equation} \label{eq:rkstep}
u^{n+1} = u^n + \dt \sum_{j  = 1}^s f(t^n + \dt c_j,y_j) b_j .
\end{equation}
Let $f_j = f(t^n + \dt c_j,y_j)$; then we can write \eqref{eq:rkstep} as
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \dt F b,
\end{equation}
where the $j$th column of $F$ is $f_j$.
We wish to impose the discrete analog of \eqref{continuous-positivity}; i.e.
\begin{align} \label{positivity}
    u^n\ge 0 \implies u^{n+1} \ge 0,
\end{align}
or more general bound constraints
\begin{align}
    \alpha \le u^n\le \beta \implies \alpha \le u^{n+1} \le \beta.
\end{align}
We will focus on the case of positivity while keeping in mind that
the methodology extends to general bounds.
The main idea of the present work is that if the new solution $u^{n+1}$ contains
negative entries, we can replace the weights in \eqref{eq:Combination} with
a set of modified weights $\bt$ such that the resulting solution is positive:
\begin{equation}\label{eq:ut}
\ut^{n+1} = u^n + \dt F \bt^n \ge 0.
\end{equation}
Indeed, we can view \eqref{eq:ut} as a linear constraint on the choice of
modified weights $\bt^n$.  Since we have already computed the intermediate stages,
$F$ is a known, fixed matrix.  In order to ensure that the modified solution $\ut^{n+1}$
is accurate, we can also constrain $\bt$ to satisfy the Runge--Kutta order conditions
up to some order (ideally, the same order as the original method).  Observe that
all of the order conditions are linear in the weights, so that these additional constraints
take the form
$$
Q\bt = r
$$
for some fixed matrix $Q$ and vector $r$.
By applying this technique at each step, we integrate \eqref{ode} with a
sequence of Runge--Kutta methods with coefficients $(A,\bt^n)$.  At any step
for which the solution $u^{n+1}$ produced by method $(A,b)$ is positive, we do not
need to modify the weights and can simply accept this unmodified solution.
Note that linear invariants (such as mass conservation) of the solution are
automatically preserved in this approach, since at each step we use a Runge--Kutta
method.

@@

To illustrate the usage of the method we consider the reaction system
\cite{kopecz_comparison_2019}
\begin{subequations}
\label{eq:Reaction}
\begin{align}
u_1' &= 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1}, \\
u_2' &= \frac{u_1u_2}{0.01+u_1}-0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2, \\
u_3' &= 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3, \\
u_4' &=0.05 u_2 + 0.02 u_3 + 0.003u_4,
\end{align}
\end{subequations}
with initial conditions
\begin{equation}
u(0) = (8,2,1,4)^T.
\end{equation}
When solved with the Cash-Karp (CK5) \cite{cash1990variable} method and $\dt = 0.005$ the approximated solution contains negative values. This causes qualitatively wrong solutions to the problem.
In Figure\,\ref{fig:exampleI} the obtained results are plotted with dashed lines.
At $t=1.905$ the value of $u_1$ gets negative. This leads to a diverging solution.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.75\textwidth]{plots/exampleI.pdf}
    \caption{Numerical approximation of the reaction problem \eqref{eq:Reaction} computed with CK5 and $\dt = 0.005$. The dashed lines are the approximations obtained without the adaption of the weights. The lower plot shows the adaptation of the weights. }
    \label{fig:exampleI}
\end{figure}

@@

The order conditions for an $s$-stage, order $p$ RKM are a set of equations depending on $A$, $b$ and
$c$.  As mentioned already, if $A$ and $c$ are given, the order conditions are
linear in $b$ and can be written as $Q_p b = r_p$, where
$Q_p\in{\mathbb R}^{v\times s}, r_p\in{\mathbb R}^v$ represent the set of all conditions up to
and including order $p$. Here $v$ is the number of order conditions. It may not be possible to find modified weights
that also satisfy the conditions of order $p$ and yield positivity, so
in general the modified weights will be a solution of
$$
  Q_\pt \bt = r_\pt
$$
for some $\pt \le p$.   Since we have $s$ degrees of freedom $\bt_j$,  we need
at a minimum to choose $\pt$ so that $\mathrm{rank}(Q_\pt) < s$.
Because the quadrature conditions are linearly independent, we have $\mathrm{rank}(Q_p)\ge p$,
so we must take $\pt \le s$.  In general we may need to take $\pt$ even smaller
in order to achieve positivity.

@@

\begin{enumerate}
    \item Select in advance a set of desirable weight vectors $b^1, b^2, \dots, b^m$
        corresponding to known good methods,
        and restrict the choice of $\bt$ to convex combinations of this set.
    \item Require that the perturbation $\|\tilde u - u\|$ is small and reject the step if it is not.
\end{enumerate}

We discuss the first strategy here; the second is deferred to section\,\ref{sec:error}.
Ideally every element of the set of potential weight vectors would correspond to
a method of the same order as the original method.   Due to linearity of the
order conditions, any linear combination of such weights would also yield a
method of the same order.  On the other hand,
it is natural to include a weight vector corresponding to the forward Euler method
(for explicit methods) or backward Euler method (for implicit methods), since
these two methods guarantee positivity (unconditionally for backward Euler
and conditionally for forward Euler).
We can formulate an LP using the first strategy (restriction to a convex set)
as follows.  Let $B$ denote the matrix with columns $b^1, b^2, \dots, b^m$ and let
$g\in{\mathbb R}^m$.  The LP is then as follows:

\LPblock{
(Convex adaptation) Given $F$, $B$, and $b$, find $g$ that minimizes $\|\bt - b\|_1$ subject to
\begin{subequations}
\begin{align}
\bt & = Bg, \\
0 & \le g_i \le 1, \\
\sum_i g_i &= 1, \\
u^{n+1}&=u^n+\dt F \bt \geq 0.
\end{align}
\end{subequations}
}

@@

\subsection{Choice of baseline method} \label{sec:integration}
An important property of the baseline method is the existence of embedded methods and the degrees of freedom for the weights $\bt$.
As noted in Section~\ref{sec:OrderCond} the number of stages has to be higher than the order.
It is natural to use explicit and diagonally implicit methods, both for their efficiency and because
the order need not be reduced as much in order to satisfy the condition $\pt<s$.
For a given method and reduced order $\pt$, the number of degrees of freedom for the choice of the new weights
is given by $s-\mathrm{rank}(Q_{\pt})$. %$\mathrm{dim}(\mathrm{ker}(Q))$.
The resulting number of degrees of freedom is shown in Table\,\ref{table:DOF_exp}
for some explicit methods and in Table\,\ref{table:DOF_imp} for several implicit
methods.

\begin{table}[h!]
\centering    %Generated below============
% \begin{tabular}{|l c|c c c c c c |}
%  \hline
% Order & &1&2&3&4&5&6 \\
%  \hline Classical RK4& s=4&3&2&0&0& - & -  \\
%  SSPRK(10,4)& s=10&9&8&6&4& - & -  \\
%  Cash-Karp RK5(4)6& s=6&5&4&2&1&0& -  \\
%  Dormand-Prince RK5(4)7& s=7&6&5&3&1&0& -  \\
%  \hline
%  \end{tabular}
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr*6c@{}}
    \toprule
    Method & $s$ & \multicolumn{6}{c}{Order $\tilde p$} \\
    & & 1 & 2 & 3 & 4 & 5 & 6 \\
    \midrule
    Classical RK4 \cite{kutta1901beitrag} & 4 & 3 & 2 & 0 & 0 & --- & --- \\
    SSPRK(10,4) \cite{ketcheson2008highly} & 10&9&8&6&4& --- & ---\\
    Cash--Karp RK5(4)6 \cite{cash1990variable} & 6&5&4&2&1&0& --- \\
    Dormand--Prince RK5(4)7 \cite{prince1981high}& 7&6&5&3&1&0& --- \\
    \bottomrule
  \end{tabular*}
  \caption{Degrees of freedom for the choice of the weights for some explicit methods.} %Generated above============
  \label{table:DOF_exp}
\end{table}

\begin{table}[h!]
\centering   %Generated below============
%   \begin{tabular}{|l c|c c c c c c |}
%   \hline
%   Order & &1&2&3&4&5&6 \\
%   \hline Implicit Euler& s=1&0& - & - & - & - & -  \\
%   Lobatto~IIIC4& s=4&3&2&1&0&0&0 \\
%   Radau~IIA3& s=3&2&1&0&0&0& -  \\
%   SDIRK(5,4)& s=5&4&3&1&0& - & -  \\
%   TR-BDF2& s=3&2&1& - & - & - & -  \\
%   Im-Euler 2& s=3&2&1& - & - & - & -  \\
%   Im-Euler~3& s=6&5&4&2& - & - & -  \\
%   Im-Euler 4& s=10&9&8&6&3& - & -  \\
%   \hline
%   \end{tabular}
   \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr*6c@{}}
    \toprule
    Method & $s$ & \multicolumn{6}{c}{Order $\tilde p$} \\
    & & 1 & 2 & 3 & 4 & 5 & 6 \\
    \midrule
    Implicit Euler& 1&0& --- & --- & --- & --- & ---  \\
    Lobatto~IIIC4 \cite{chipman1971stable} & 4&3&2&1&0&0&0 \\
    Radau~IIA3 \cite{ehle1969pade} & 3&2&1&0&0&0& ---  \\
    SDIRK(5,4) \cite[eq. (6.18)]{hairer_solving_1996}& 5&4&3&1&0& --- & ---  \\
    TR-BDF2 \cite{bank1985transient} & 3&2&1& --- & --- & --- & ---  \\
    Extrapolation Im-Euler 2& 3&2&1& --- & --- & --- & ---  \\
    Extrapolation Im-Euler~3& 6&5&4&2& --- & --- & ---  \\
    Extrapolation Im-Euler 4& 10&9&8&6&3& --- & ---  \\
    \bottomrule
  \end{tabular*}
  \caption{Degrees of freedom for the choice of the weights for some implicit methods.} %Generated above============
  \label{table:DOF_imp}
\end{table}

@@

Regarding implicit methods,
we can see that the fully implicit methods Lobatto~IIIC4 and Radau~IIA3 require a drastic reduction of the order, as expected.
The diagonally implicit SDIRK(5,4) method only requires an order reduction of one to get one degree of freedom for the weights.
The TR-BDF2 method even allows adaptations without reducing the order.
The implicit Euler extrapolation methods also exhibit degrees of freedom without a reduction of the order.

@@

To approximate the error of a new step we propose the following approximation of the local error:
\begin{align}
err = \|u(t^{n+1})-\tilde u^{n+1}\| &= \|u(t^{n+1}) - (u^{n+1}+\dt F(\bt-b))\| \\
 &\leq \underbrace{\|u(t^{n+1})-u^{n+1}\|}_{\approx err_T}+\underbrace{\|\dt F(\bt-b)\|}_{= \delta} \label{eq:Err}
\end{align}

@@

As an example, the stability regions of adapted RKMs are visualized in
Figure\,\ref{fig:stab}.
In Figure\,\ref{fig:stab_dp5} the Dormand--Prince RK5 method is freely adapted.
The weights are taken from the example in Section~\ref{sec:Ex_expl}.
In Figure\,\ref{fig:stab_ex3} the stability regions of the Im-Euler~3
extrapolation method and the embedded chain of three BE steps are
plotted. Additionally the stability regions of convex combinations
of these two methods are shown.

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{plots/stab_dp5.pdf}
         \caption{Dormand--Prince RK5 with free adaptation
                   of the weights $\bt$ as in the example in Figure~\ref{fig:weights_AdDe}.}
         \label{fig:stab_dp5}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{plots/stab_ex3.pdf}
         \caption{Im-Euler~3 extrapolation method and embedded chain
                  of three BE steps with convex combinations of both
                  weights.}
         \label{fig:stab_ex3}
     \end{subfigure}
        \caption{Change of stability region for the free adaptation and
                 the convex adaptation of the weights.
                 Regions with $|R(z)| \leq 1$ are hatched for the original methods.}
        \label{fig:stab}
\end{figure}

@@

Next we test the adaptive RKM on a more complex problem.
For this, we consider the advection-diffusion-production-destruction system  \cite{kopecz_comparison_2019}
\begin{subequations}
\label{eq:ADR}
\begin{align}
\frac{\partial u_1}{\partial t} &=-a \frac{\partial u_1}{\partial x} + d\frac{\partial^2 u_1}{\partial x ^2} + 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1}, \\
\frac{\partial u_2}{\partial t} &=-a \frac{\partial u_2}{\partial x} + d\frac{\partial^2 u_2}{\partial x ^2} + \frac{u_1u_2}{0.01+u_1} -0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2, \\
\frac{\partial u_3}{\partial t} &=-a \frac{\partial u_3}{\partial x} + d\frac{\partial^2 u_3}{\partial x ^2} + 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3, \\
\frac{\partial u_4}{\partial t} &=-a \frac{\partial u_4}{\partial x} + d\frac{\partial^2 u_4}{\partial x ^2} + 0.05 u_2 + 0.02 u_3 - 0.003u_4,
\end{align}
\end{subequations}
with parameters $a=\num{1e-2} $ and $ d=\num{1e-6}$.
The PDE is simulated on the domain $x = [0,1]$ with $N=100$ points and periodic boundary conditions.
The advection part is semidiscretized using a first order upwind scheme and the diffusion part is semidiscretized using a central 3-point-scheme. This leads to a positivity preserving system of ODEs which conserves the total mass $\sum u$.
The computation is done using the Im-Euler~3 extrapolation
method with free adaptation.
As step size control a PI-control from\,\cite{hairer_solving_1996} is used. The error was estimated using \eqref{eq:Err}. The tolerance was set to $Tol = 0.01$.
The final time is $t_{end} = 50$.

>>>>

We say an initial value problem
\begin{align} \label{ode}
    u'(t) & = f(t,u) \\
    u(0) & = u_0
\end{align}
where $u\colon [0,T] \to \R^m$ is positive if
\begin{align} \label{continuous-positivity}
    u(0)\ge 0 \implies u(t) \ge 0 \text{ for all } t \in [0,T].
\end{align}
Here and in the following, inequalities like $u \geq 0$ are meant componentwise.
A sufficient condition for positivity of \eqref{ode} is
\begin{equation} \label{eq:condition_ODE_pos}
u_i=0 \implies f_i(t,u_1,\cdots,u_i,\cdots,u_n) \geq 0 \quad \forall {u \geq 0,}\; \ \  \forall {t\in[0,T]}.
\end{equation}
For such ODEs, the backward Euler method is guaranteed to preserve positivity
under any step size, while the forward Euler will preserve positivity for small enough $\dt$ \cite{hundsdorfer_numerical_2003}.
Any RKM (or in fact any general linear method) that is unconditionally positivity preserving for all positive ODEs
must have order $\le 1$ \cite{bolley_conservation_1978}.
For any higher order method, we expect positivity only under some restriction
on the time step size.

@@

The idea of using different weights within an RKM is not new; for instance it is
the basis of error approximation using embedded RK pairs \cite{hairer_solving_1993}.
The idea of adapting the weights after calculating the stage values has also been used,
for instance in \cite{ketcheson_spatially_2013}.  In this case it is used to
adapt the properties of the time integrator for a method of lines solution of a
PDE.  Another class of methods that adapt the weights at the end of an RK step
are the relaxation Runge--Kutta (RRK) methods.  In these, the weights are
scaled by a scalar relaxation parameter in order to guarantee conservation or
monotonicity of a desired functional; e.g.\ to conserve or dissipate energy or
entropy
\cite{ketcheson_relaxation_2019,ranocha_relaxation_2019,ranocha2020general}.

Our means to ensure positivity can be interpreted as a projection
approach, where the numerical solution is adapted to satisfy the
positivity constraint at the end of each step. In contrast to
simple orthogonal projection, which has also been proposed to deal
with positivity constraints \cite{shampine1986conservation},
our approach preserves all linear invariants of the given ODE. These
invariants can be very important, e.g.\ the total mass for a transport
problem or in reaction systems. Preservation of linear
invariants has been shown to be an important advantage of RRK methods
over orthogonal projection methods
\cite{ranocha2020relaxationHamiltonian}. Of course, it is also
possible to enforce the preservation of linear invariants in projection
methods, but the invariants have to be known explicitly
\cite{sandu2001positive}.

@@

When computing the solution of an ODE $u ' = f(t,u) $ using an RKM with $s$ stages and the Butcher tableau
\begin{align}
\renewcommand{\arraystretch}{1.2}
\begin{array}{c|c}
c &  A \\
\hline
 & b^T\\
\end{array}
\end{align}
the stage values are computed according to
\begin{equation}\label{eq:stagevalues}
y_j =  u^n + \dt \sum_{k = 1}^{s} a_{jk} f(t^n + \dt c_k,y_k),  \quad j = 1,\cdots,s.
\end{equation}
Based on these values, the next solution $u^{n+1}$ is computed as
\begin{equation} \label{eq:rkstep}
u^{n+1} = u^n + \dt \sum_{j  = 1}^s f(t^n + \dt c_j,y_j) b_j .
\end{equation}
Let $f_j = f(t^n + \dt c_j,y_j)$; then we can write \eqref{eq:rkstep} as
\begin{equation}\label{eq:Combination}
u^{n+1} = u^n + \dt F b,
\end{equation}
where the $j$th column of $F$ is $f_j$.
We wish to impose the discrete analog of \eqref{continuous-positivity}; i.e.\
\begin{align} \label{positivity}
    u^n\ge 0 \implies u^{n+1} \ge 0,
\end{align}
or more general bound constraints
\begin{align}
    \alpha \le u^n\le \beta \implies \alpha \le u^{n+1} \le \beta.
\end{align}
We will focus on the case of positivity while keeping in mind that
the methodology extends to general bounds.
The main idea of the present work is that if the new solution $u^{n+1}$ contains
negative entries, we can replace the weights in \eqref{eq:Combination} with
a set of modified weights $\bt$ such that the resulting solution is positive:
\begin{equation}\label{eq:ut}
\ut^{n+1} = u^n + \dt F \bt^n \ge 0.
\end{equation}
Indeed, we can view \eqref{eq:ut} as a linear constraint on the choice of
modified weights $\bt^n$.  Since we have already computed the intermediate stages,
$F$ is a known, fixed matrix.  In order to ensure that the modified solution $\ut^{n+1}$
is accurate, we can also constrain $\bt$ to satisfy the Runge--Kutta order conditions
up to some order (ideally, the same order as the original method).  Observe that
all of the order conditions are linear in the weights, so that these additional constraints
take the form
$$
Q\bt = r
$$
for some fixed matrix $Q$ and vector $r$.
By applying this technique at each step, we integrate \eqref{ode} with a
sequence of Runge--Kutta methods with coefficients $(A,\bt^n)$.  At any step
for which the solution $u^{n+1}$ produced by method $(A,b)$ is positive, we do not
need to modify the weights and can simply accept this unmodified solution.
Note that linear invariants (such as mass conservation) of the solution are
automatically preserved in this approach, since at each step we use a Runge--Kutta
method.

@@

To illustrate the usage of the method we consider the reaction system
\cite{kopecz_comparison_2019}
\begin{subequations}
\label{eq:Reaction}
\begin{align}
u_1' &= 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1}, \\
u_2' &= \frac{u_1u_2}{0.01+u_1}-0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2, \\
u_3' &= 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3, \\
u_4' &=0.05 u_2 + 0.02 u_3 + 0.003u_4,
\end{align}
\end{subequations}
with initial conditions
\begin{equation}
u(0) = (8,2,1,4)^T.
\end{equation}
When solved with the Cash--Karp RK5 method \cite{cash1990variable} and $\dt = 0.005$ the approximated solution contains negative values. This causes qualitatively wrong solutions to the problem.
In Figure\,\ref{fig:exampleI} the obtained results are plotted with dashed lines.
At $t=1.905$ the value of $u_1$ gets negative. This leads to a diverging solution.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.75\textwidth]{plots/exampleI.pdf}
    \caption{Numerical approximation of the reaction problem \eqref{eq:Reaction} computed with Cash--Karp RK5 and $\dt = 0.005$. The dashed lines are the approximations obtained without the adaption of the weights. The lower plot shows the adaptation of the weights. }
    \label{fig:exampleI}
\end{figure}

@@

The order conditions for an $s$-stage, order $p$ RKM are a set of equations depending on $A$, $b$, and
$c$.  As mentioned already, if $A$ and $c$ are given, the order conditions are
linear in $b$ and can be written as $Q_p b = r_p$, where
$Q_p\in{\mathbb R}^{v\times s}, r_p\in{\mathbb R}^v$ represent the set of all conditions up to
and including order $p$. Here $v$ is the number of order conditions. It may not be possible to find modified weights
that also satisfy the conditions of order $p$ and yield positivity, so
in general the modified weights will be a solution of
$$
  Q_\pt \bt = r_\pt
$$
for some $\pt \le p$.   Since we have $s$ degrees of freedom $\bt_j$,  we need
at a minimum to choose $\pt$ so that $\mathrm{rank}(Q_\pt) < s$.
Because the quadrature conditions are linearly independent, we have $\mathrm{rank}(Q_p)\ge p$,
so we must take $\pt \le s$.  In general we may need to take $\pt$ even smaller
in order to achieve positivity.

@@

\begin{enumerate}
    \item Select in advance a set of desirable weight vectors $b^1, b^2, \dots, b^K$
        corresponding to known good methods,
        and restrict the choice of $\bt$ to convex combinations of this set.
    \item Require that the perturbation $\|\tilde u - u\|$ is small and reject the step if it is not.
\end{enumerate}

We discuss the first strategy here; the second is deferred to section\,\ref{sec:error}.
Ideally every element of the set of potential weight vectors would correspond to
a method of the same order as the original method.   Due to linearity of the
order conditions, any linear combination of such weights would also yield a
method of the same order.  On the other hand,
it is natural to include a weight vector corresponding to the forward Euler method
(for explicit methods) or backward Euler method (for implicit methods), since
these two methods guarantee positivity (unconditionally for backward Euler
and conditionally for forward Euler).
We can formulate an LP using the first strategy (restriction to a convex set)
as follows.  Let $B$ denote the matrix with columns $b^1, b^2, \dots, b^K$ and let
$g\in \R^K$.  The LP is then as follows:

\LPblock{
(Convex adaptation) Given $F$, $B$, and $b$, find $g$ that minimizes $\|\bt - b\|_1$ subject to
\begin{subequations}
\begin{align}
\bt & = Bg, \\
0 & \le g_k \le 1, \\
\sum_{k=1}^K g_k &= 1, \\
u^{n+1}&=u^n+\dt F \bt \geq 0.
\end{align}
\end{subequations}
}

@@

\subsection{Choice of baseline method} \label{sec:integration}
An important property of the baseline method is the existence of embedded methods and the degrees of freedom for the weights $\bt$.
As noted in Section~\ref{sec:OrderCond} the number of stages has to be higher than the order.
It is natural to use explicit and diagonally implicit methods, both for their efficiency and because
the order need not be reduced as much in order to satisfy the condition $\pt<s$.
For a given method and reduced order $\pt$, the number of degrees of freedom for the choice of the new weights
is given by $s-\mathrm{rank}(Q_{\pt})$. %$\mathrm{dim}(\mathrm{ker}(Q))$.
The resulting number of degrees of freedom is shown in Table\,\ref{table:DOF_exp}
for some explicit methods and in Table\,\ref{table:DOF_imp} for several implicit
methods. The backward Euler extrapolation methods use the harmonic sequence
as described in \cite[Section~II.9]{hairer_solving_1993} and
\cite[Section~IV.9]{hairer_solving_1996}.

\begin{table}[h!]
\centering    %Generated below============
% \begin{tabular}{|l c|c c c c c c |}
%  \hline
% Order & &1&2&3&4&5&6 \\
%  \hline Classical RK4& s=4&3&2&0&0& - & -  \\
%  SSPRK(10,4)& s=10&9&8&6&4& - & -  \\
%  Cash--Karp RK5(4)6& s=6&5&4&2&1&0& -  \\
%  Dormand-Prince RK5(4)7& s=7&6&5&3&1&0& -  \\
%  \hline
%  \end{tabular}
  \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr*6c@{}}
    \toprule
    Method & $s$ & \multicolumn{6}{c}{Order $\tilde p$} \\
    & & 1 & 2 & 3 & 4 & 5 & 6 \\
    \midrule
    Classical RK4 \cite{kutta1901beitrag} & 4 & 3 & 2 & 0 & 0 & --- & --- \\
    SSPRK(10,4) \cite{ketcheson2008highly} & 10&9&8&6&4& --- & ---\\
    Cash--Karp RK5(4)6 \cite{cash1990variable} & 6&5&4&2&1&0& --- \\
    Dormand--Prince RK5(4)7 \cite{prince1981high}& 7&6&5&3&1&0& --- \\
    \bottomrule
  \end{tabular*}
  \caption{Degrees of freedom for the choice of the weights for some explicit methods.} %Generated above============
  \label{table:DOF_exp}
\end{table}

\begin{table}[h!]
\centering   %Generated below============
%   \begin{tabular}{|l c|c c c c c c |}
%   \hline
%   Order & &1&2&3&4&5&6 \\
%   \hline Implicit Euler& s=1&0& - & - & - & - & -  \\
%   Lobatto~IIIC4& s=4&3&2&1&0&0&0 \\
%   Radau~IIA3& s=3&2&1&0&0&0& -  \\
%   SDIRK(5,4)& s=5&4&3&1&0& - & -  \\
%   TR-BDF2& s=3&2&1& - & - & - & -  \\
%   Im-Euler 2& s=3&2&1& - & - & - & -  \\
%   Im-Euler 3& s=6&5&4&2& - & - & -  \\
%   Im-Euler 4& s=10&9&8&6&3& - & -  \\
%   \hline
%   \end{tabular}
   \begin{tabular*}{\linewidth}{@{\extracolsep{\fill}}lr*6c@{}}
    \toprule
    Method & $s$ & \multicolumn{6}{c}{Order $\tilde p$} \\
    & & 1 & 2 & 3 & 4 & 5 & 6 \\
    \midrule
    Backward Euler& 1&0& --- & --- & --- & --- & ---  \\
    Lobatto~IIIC4 \cite{chipman1971stable} & 4&3&2&1&0&0&0 \\
    Radau~IIA3 \cite{ehle1969pade} & 3&2&1&0&0&0& ---  \\
    SDIRK(5,4) \cite[eq. (6.18)]{hairer_solving_1996}& 5&4&3&1&0& --- & ---  \\
    TR-BDF2 \cite{bank1985transient} & 3&2&1& --- & --- & --- & ---  \\
    Extrapolation BE~2 \cite[Sec.~II.9]{hairer_solving_1993} & 3&2&1& --- & --- & --- & ---  \\
    Extrapolation BE~3 \cite[Sec.~II.9]{hairer_solving_1993} & 6&5&4&2& --- & --- & ---  \\
    Extrapolation BE~4 \cite[Sec.~II.9]{hairer_solving_1993} & 10&9&8&6&3& --- & ---  \\
    \bottomrule
  \end{tabular*}
  \caption{Degrees of freedom for the choice of the weights for some implicit methods.} %Generated above============
  \label{table:DOF_imp}
\end{table}

@@

Regarding implicit methods,
we can see that the fully implicit methods Lobatto~IIIC4 and Radau~IIA3 require a drastic reduction of the order, as expected.
The diagonally implicit SDIRK(5,4) method only requires an order reduction of one to get one degree of freedom for the weights.
The TR-BDF2 method even allows adaptations without reducing the order.
The backward Euler extrapolation methods also exhibit degrees of freedom without a reduction of the order.

@@

To approximate the error of a new step we propose the following approximation of the local error:
\begin{align}
err = \|u(t^{n+1})-\tilde u^{n+1}\| &= \|u(t^{n+1}) - (u^{n+1}+\dt F(\bt-b))\| \\
 &\leq \underbrace{\|u(t^{n+1})-u^{n+1}\|}_{\approx err_T}+\underbrace{\|\dt F(\bt-b)\|}_{= \delta}. \label{eq:Err}
\end{align}

@@

As an example, the stability regions of adapted RKMs are visualized in
Figure\,\ref{fig:stab}.
In Figure\,\ref{fig:stab_dp5} the Dormand--Prince RK5 method is freely adapted.
The weights are taken from the example in Section~\ref{sec:Ex_expl}.
In Figure\,\ref{fig:stab_ex3} the stability regions of the BE~3
extrapolation method and the embedded chain of three BE steps with
time step $\dt/3$ are
plotted. Additionally the stability regions of convex combinations
of these two methods are shown.

\begin{figure}
     \centering
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{plots/stab_dp5.pdf}
         \caption{Dormand--Prince RK5 with free adaptation
                   of the weights $\bt$ as in the example in Figure~\ref{fig:weights_AdDe}.}
         \label{fig:stab_dp5}
     \end{subfigure}
     \hfill
     \begin{subfigure}[b]{0.45\textwidth}
         \centering
         \includegraphics[width=\textwidth]{plots/stab_ex3.pdf}
         \caption{BE~3 extrapolation method and embedded chain
                  of three BE steps with convex combinations of both
                  weights.}
         \label{fig:stab_ex3}
     \end{subfigure}
        \caption{Change of stability region for the free adaptation and
                 the convex adaptation of the weights.
                 Regions with $|R(z)| \leq 1$ are hatched for the original methods.}
        \label{fig:stab}
\end{figure}

@@

Next we test the adaptive RKM on a more complex problem.
For this, we consider the advection-diffusion-production-destruction system  \cite{kopecz_comparison_2019}
\begin{subequations}
\label{eq:ADR}
\begin{align}
\frac{\partial u_1}{\partial t} &=-a \frac{\partial u_1}{\partial x} + d\frac{\partial^2 u_1}{\partial x ^2} + 0.01u_2 + 0.01 u_3 +0.003u_4 - \frac{u_1 u_2}{0.01+u_1}, \\
\frac{\partial u_2}{\partial t} &=-a \frac{\partial u_2}{\partial x} + d\frac{\partial^2 u_2}{\partial x ^2} + \frac{u_1u_2}{0.01+u_1} -0.01 u_2-0.5(1-\exp(-1.21 u_2^2)) u_3 -0.05 u_2, \\
\frac{\partial u_3}{\partial t} &=-a \frac{\partial u_3}{\partial x} + d\frac{\partial^2 u_3}{\partial x ^2} + 0.5(1-\exp(-1.21u_2^2)) u_3 - 0.01 u_3 -0.02 u_3, \\
\frac{\partial u_4}{\partial t} &=-a \frac{\partial u_4}{\partial x} + d\frac{\partial^2 u_4}{\partial x ^2} + 0.05 u_2 + 0.02 u_3 - 0.003u_4,
\end{align}
\end{subequations}
with parameters $a=\num{1e-2} $ and $ d=\num{1e-6}$.
The PDE is simulated on the domain $x = [0,1]$ with $N=100$ points and periodic boundary conditions.
The advection part is semidiscretized using a first order upwind scheme and the diffusion part is semidiscretized using a central 3-point-scheme. This leads to a positivity preserving system of ODEs which conserves the total mass $\sum u$.
The computation is done using the BE~3 extrapolation
method with free adaptation.
As step size control a PI-control from\,\cite{hairer_solving_1996} is used. The error was estimated using \eqref{eq:Err}. The tolerance was set to $Tol = 0.01$.
The final time is $t_{end} = 50$.
